### 2014-07-05 paper.tex
minor changes

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Comparison of spliced reads and reference gene annotations has been successfully used to discover alternative
splicing in model organisms. However, the method cannot be applied in organisms without high-quality reference genome and
gene annotations. In this article, we introduce a pipeline, based on \emph{de novo} assembly, that constructs gene models with splice
variants.
The pipeline includes a new technique called local assembly that enhances the sensitivity of
alternative splicing detection.
We demonstrates that the pipeline detects many novel splice variants in RNA-Seq data from chicken spleens.
Some of those splice variants; however, only detected by our pipeline and not Cufflinks.
This indicates that the pipeline can be used to facilitate splice variants detection from RNA-Seq.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\section*{Author Summary}

\section*{Introduction}
Until recently, studies of alternative splicing had been limited to a small
number of genes and isoforms due to high-cost and low-throughput sequencing
of expressed sequence tags (ESTs) and full-length cDNA libraries.  
RNA sequencing (RNA-Seq) using the next-generation sequencing (NGS) has been
used successfully in many studies to gain unprecedented insight into a
complexity of transcriptomes.
It has been estimated that, in human, $92-94$\% of multiexon genes undergo alternative
splicing and different isoforms are expressed in different tissues\cite{Wang:2008ea}.
This suggests that even in human a large number of splice variants have not been explored.

@@

Despite the small size of sequencing reads, several studies have detected
novel splice junctions based on alignment of reads spanning across putative exon junctions.
To map reads across exon junctions, reads are split into two parts and each part is mapped
to the genome independently.
A splice junction is then identified based on alignments of each half of a read
that fall between two exons at exon-intron boundaries.
With this approach, Wang \emph{et al} have identified a large number of
splice junctions that are not annotated from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.
These novel splice sites include both canonical and non-canonical splice sites.
Approximately, $46\%-75\%$ of canonical splice sites are supported by ESTs.
Novel splice junctions have different levels of read coverage suggesting that both high-
and low-expressed isoforms are unannotated.

Using a similar approach, Pickrell \emph{et al} identified more than 150,000 novel canonical splice junctions
in lymphoblastoid cells.
The study also shows that the number of unannotated splice junctions varies among cells from
different human tissues, which suggests tissue-specific expression of isoforms\cite{Pickrell:2010gt}.
A majority of unannotated isoforms found in both studies are from alternative splicing
of annotated exons. Only a small fraction of isoforms contain exons from intergenic or
intronic region.

Several tools have been developed not only to detect novel splice junctions but also to reconstruct
full-length isoforms from short reads without using gene annotations.
These tools are especially useful for transcriptome analysis of organims with poor gene annotations.
Cufflinks\cite{Trapnell:2010kd} relies on splice junctions detected from Tophat\cite{Trapnell:2009dp},
a read aligner that can align reads across putative exon junctions, to reconstruct a full-length transcript.
Cufflinks identified 12,712 novel isoforms, of which 7,395 ($58\%$) contain novel splice junctions in
mouse myoblast cell lines.
Guttman \emph{et al} used Scripture, a tool employing similar mapping-based approach, to reconstruct
a full-length transcripts from mouse RNA-Seq data and discovered approximately 490 novel alternative
isoforms in lincRNA loci, which are expressed in any of the three different cell types\cite{Guttman:2010io}.

Although the mapping-based method has been useful in both detecting splice junctions and isoforms,
it relies heavily on a reference genome.
Hence, organisms lacking a genome or a high-quality one may not find this approach practical.
This limitation can be overcome by \emph{de novo} assembly of short reads.

A number of \emph{de novo} assemblers have been used to reconstruct transcripts from RNA-Seq data in
many studies.
Trinity\cite{Grabherr:2011jb}, was successfully used to reconstruct transcripts from yeast and mouse datasets.
It was also shown that Trinity detects a unique set of novel splice junctions not detected by Cufflinks
or Scripture.
This suggests that \emph{de novo} assembly approach is capabale of increasing sensitivity of detecting
alternative isoforms of a mapping-based method.
Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are an extension
of Abyss\cite{Simpson:2009iv} and Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} genome assembler
that are tuned to work with RNA-Seq data.
These assemblers are comparable at reconstructing existing and novel alternative isoforms with a slightly
different sensitivity and specificity.
However, Oases with Oases-M has been shown to be superior than other \emph{de novo}
assemblers at discovering isoforms in human and mouse\cite{Schulz:2012je}.

In this study, we present a pipeline that uses Velvet and Oases assembler to
reconstruct alternative isoforms in RNA-Seq data from chickens.
We use a new technique called a local assembly that enhances sensitivity of
alternative isoform detection of Oases.
The results show that the pipeline can detect more isoforms than Oases-M and
can detect isoforms not found by Cufflinks.
We also showed that transcripts reconstructed from \emph{de novo} assembly and mapping-based approach can be
merged to build more complete predicted gene models.

@@

We next used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While the local assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly
% @CTB why are the global ones not included in the final gene models?
(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
the incomplete nature of the current chicken genome assembly, which is
lacking approximately 5\% of its true gene content (cite).
Interestingly, over a hundred regions were present {\em only} in the
local assemblies, suggesting that the local assemblies might be
recovering additional exons.  Significant numbers of unique regions
from both global and local assemblies showed homology to the mouse
genome, indicating that at least some of these unique sequences
represented real sequence content.  Figure~\ref{global_vs_local} shows
an example of different isoforms detected by the two assembly methods.

@@

% @CTB DISCSUSS the use of splice junctions for validity; transition
% from transcripts to splice junctions.

@@

\subsection*{Exon graphs can be used to reconstruct putative splice variants}

We developed an exon graph approach to construct gene models from
alignments of transcripts against the genome.  Our approach,
implemented in a software package called Gimme, merges transcripts and
gene models based on overlapping exons using an exon-graph approach
(see Materials and Methods; Figure~\ref{algorithm}).  We used Gimme to obtain
20,439 gene models containing 25,859 isoforms from our global
assembly; 21,963 gene models containing 25,859 from our local
assembly; and 20,922 gene models containing 27,732 isoforms from the
merged global and local assemblies (Table~\ref{genes_transcripts}).

@@

We used bowtie to map single-end reads from the source datasets to the
transcripts.  More than 80\% of the original reads could be aligned to
the transcripts, demonstrating that we did not lose a significant
amount of information during the merge process
(Table~\ref{single-end_map}).  More importantly, we also mapped
paired-end reads from technical replicates to the same gene models,
and found that more than 85\% of the paired-ends mapped concordantly
the gene models (Table~\ref{paired-end_map}).  Most of the reads that
did not map were either highly erroneous or contained low-complexity
artifactual sequence that probably originated from sample processing
and reverse transcription (cite).
% @CTB do we have something we can cite for typical % mapping?  Maybe Trinity
% paper?  How about a citation that suggests that there's a lot of crap
% in mRNAseq?
Thus the merged gene models produced by Gimme represent the significant
majority of the assemblable data.

@@

Of 100,411 splice junctions from the gene models, 95 (0.09\%)
junctions have no spliced reads and only 448 (0.45\%) junctions
have fewer than 4 spliced reads (Fig.~\ref{cdf_single_splice}).
% @CTB how do you get junctions with no spliced reads!?
More than 99\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.
% @CTB I'm slightly bugged by this statement.  Are we right?

@@

We next compared the Gimme gene models to those produced by Cufflinks,
another reference-based approach to building gene models from mRNAseq
data\cite{Trapnell:2010kd}.  We also compared the results from both methods to the ENSEMBL gene annotations,
which are produced by a pipeline that incorporates de novo gene
prediction and homology-based approaches as well as expression data.

@@

The ENSEMBL pipeline does not, however, include a large number of
splice junctions from ESTs (99,755) or mRNAs (13,641).  Cufflinks and
Gimme each recover about 10\% of these, with more than 2/3 of these
recovered by both Cufflinks and Gimme.  This indicates that both Gimme
and Cufflinks are equally adept at recovering novel splice junctions.
% @CTB note this suggests that Cufflinks improved recovery for ENSEMBL
% gene models is false...
% @CTB what fraction of Cufflinks splice junctions are high coverage (4+)?

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Comparison of spliced reads and reference gene annotations has been
successfully used to discover alternative splicing in model
organisms. However, the method cannot be applied in organisms without
high-quality reference genome and gene annotations. In this article,
we introduce a pipeline, based on \emph{de novo} assembly, that
constructs gene models with splice variants.  The pipeline uses a
technique called local assembly that enhances the sensitivity of
alternative splicing detection.  We demonstrate that the pipeline
detects many novel splice variants in RNA-Seq data from chicken
spleen.  Many of these splice variants; however, are only detected by our
pipeline and not Cufflinks.  This indicates that the pipeline can be
used to facilitate splice variant detection from RNA-Seq.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.

% @CTB: aren't some splice variants detected only by Cufflinks and not by
% gimme, too?
% @CTB: for me to check, do Tophat and cufflinks only detect canonical splice?
% @CTB: ask Olga for references to single-cell splice stuff.

\section*{Author Summary}

\section*{Introduction}

Until recently, studies of alternative splicing have been limited to a
small number of genes and isoforms due to high-cost and low-throughput
sequencing of expressed sequence tags (ESTs) and full-length cDNA
libraries.  RNA sequencing (RNA-Seq) using deep short-read sequencing
has been used successfully in many studies to gain unprecedented
insight into a complexity of transcriptomes.

It has been estimated that, in human, $92-94$\% of multiexon genes
undergo alternative splicing and different isoforms are expressed in
different tissues\cite{Wang:2008ea}.  This suggests that even in human
a large number of splice variants have not been explored.

@@

Despite the small size of sequencing reads, several studies have
detected novel splice junctions based on alignment of reads spanning
across putative exon junctions.  To map reads across exon junctions,
reads are split into two parts and each part is mapped to the genome
independently (@cite).  A splice junction is then identified based on
alignments of each half of a read that fall between two exons at
exon-intron boundaries.  With this approach, Wang \emph{et al} have
identified a large number of splice junctions that are not annotated
from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.  These novel
splice sites include both canonical and non-canonical splice sites.
Approximately, $46\%-75\%$ of canonical splice sites are supported by
ESTs.  Novel splice junctions have different levels of read coverage
suggesting that both high- and low-expressed isoforms are unannotated.

Using a similar approach, Pickrell \emph{et al} identified more than
150,000 novel canonical splice junctions in lymphoblastoid cells.  The
study also shows that the number of unannotated splice junctions
varies among cells from different human tissues, which suggests
tissue-specific expression of isoforms\cite{Pickrell:2010gt}.  A
majority of unannotated isoforms found in both studies are from
alternative splicing of annotated exons. Only a small fraction of
isoforms contain previously unknown exons.
% @CTB is this last point important?

Several tools have been developed not only to detect novel splice
junctions but also to reconstruct full-length isoforms from short
reads without using prior gene annotations.  These tools are especially
useful for transcriptome analysis of organims with incomplete gene
annotations.  Cufflinks\cite{Trapnell:2010kd} relies on splice
junctions detected from Tophat\cite{Trapnell:2009dp}, a read aligner
that can align reads across putative exon junctions, to reconstruct a
full-length transcript.  Cufflinks identified 12,712 novel isoforms,
of which 7,395 ($58\%$) contain novel splice junctions in mouse
myoblast cell lines.  Guttman \emph{et al} used Scripture, a tool
employing similar mapping-based approach, to reconstruct a full-length
transcripts from mouse RNA-Seq data and discovered approximately 490
novel alternative isoforms in lincRNA loci, which are expressed in any
of the three different cell types\cite{Guttman:2010io}.

Although these mapping-based methods have been useful in detecting
both splice junctions and isoforms, they rely heavily on a reference
genome.  Hence, it is not necessarily practical to apply these methods
to organisms lacking a high-quality reference genome.  This limitation
can be overcome by \emph{de novo} assembly of short reads.

A number of \emph{de novo} assemblers have been used to reconstruct
transcripts from RNA-Seq data in many studies.
Trinity\cite{Grabherr:2011jb}, was successfully used to reconstruct
transcripts from yeast and mouse datasets.  It was also shown that
Trinity detects a unique set of novel splice junctions not detected by
Cufflinks or Scripture.  This suggests that a \emph{de novo} assembly
approach is capabale of increasing sensitivity of detecting
alternative isoforms of a mapping-based method.
Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are
extensions of the Abyss\cite{Simpson:2009iv} and
Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} genome assemblers that are
tuned to work with RNA-Seq data.  These assemblers are comparable at
reconstructing existing and novel alternative isoforms with a slightly
different sensitivity and specificity.  However, Oases with Oases-M
has been shown to be superior than other \emph{de novo} assemblers at
discovering isoforms in human and mouse\cite{Schulz:2012je}.

In this study, we present a pipeline that uses {\em de novo} assembly
to reconstruct alternative isoforms in RNA-Seq data from chickens.  We
apply a technique we call ``local assembly'' that enhances the sensitivity
of alternative isoform detection by Oases.  The results show that the
pipeline can detect more isoforms than Oases-M and can detect isoforms
not found by Cufflinks.  We also showed that transcripts reconstructed
from \emph{de novo} assembly and mapping-based approach can be merged
to build more complete gene models.

@@

We next used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While the local assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly
% @CTB why are the global ones not included in the final gene models?
(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
the incomplete nature of the current chicken genome assembly, which is
lacking approximately 5\% of its true gene content (@cite).
Interestingly, over a hundred regions were present {\em only} in the
local assemblies, suggesting that the local assemblies might be
recovering additional exons.  Significant numbers of unique regions
from both global and local assemblies showed homology to the mouse
genome, indicating that at least some of these unique sequences
represented real sequence content.  Figure~\ref{global_vs_local} shows
an example of different isoforms detected by the two assembly methods.

@@

% @CTB DISCUSS the use of splice junctions for validity; transition
% from transcripts to splice junctions.

@@

\subsection*{Exon graphs can reconstruct putative splice variants}

We used an exon graph approach to construct gene models from
alignments of transcripts against the genome.  Our approach,
implemented in a software package called Gimme, merges transcripts and
gene models based on overlapping exons using an exon-graph approach
(see Materials and Methods; Figure~\ref{algorithm}).  We used Gimme to obtain
20,439 gene models containing 25,859 isoforms from our global
assembly; 21,963 gene models containing 25,859 from our local
assembly; and 20,922 gene models containing 27,732 isoforms from the
merged global and local assemblies (Table~\ref{genes_transcripts}).

@@

We used bowtie to map single-end reads from the source datasets to the
transcripts.  More than 80\% of the original reads could be aligned to
the transcripts, demonstrating that we did not lose a significant
amount of information during the merge process
(Table~\ref{single-end_map}).
% @CTB compared to what?
More importantly, we also mapped
paired-end reads from technical replicates to the same gene models,
and found that more than 85\% of the paired-ends mapped concordantly
the gene models (Table~\ref{paired-end_map}).  Most of the reads that
did not map were either highly erroneous or contained low-complexity
artifactual sequence that probably originated from sample processing
and reverse transcription (@cite).
% @CTB do we have something we can cite for typical % mapping?  Maybe Trinity
% paper?  How about a citation that suggests that there's a lot of crap
% in mRNAseq?
Thus the merged gene models produced by Gimme represent the significant
majority of the assemblable data.

@@

Of 100,411 splice junctions from the gene models, 95 (0.09\%)
junctions have no spliced reads and only 448 (0.45\%) junctions
have fewer than 4 spliced reads (Fig.~\ref{cdf_single_splice}).
% @CTB how do you get junctions with no spliced reads!?
More than 99\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.
% @CTB I'm slightly bugged by this statement.  Are we right?!

@@

We next compared the Gimme gene models to those produced by Cufflinks,
another reference-based approach to building gene models from mRNAseq
data\cite{Trapnell:2010kd}.  We also compared the results from both
methods to the ENSEMBL gene annotations, which are produced by a
pipeline that incorporates de novo gene prediction and homology-based
approaches as well as expression data (@cite).

@@

The ENSEMBL pipeline does not, however, include a large number of
splice junctions from ESTs (99,755) or mRNAs (13,641).  Cufflinks and
Gimme each recover about 10\% of these, with more than 2/3 of these
recovered by both Cufflinks and Gimme.  This indicates that both Gimme
and Cufflinks are equally adept at recovering novel splice junctions.

% @CTB note this suggests that Cufflinks improved recovery for ENSEMBL
% gene models is false...
% @CTB what fraction of Cufflinks splice junctions are high coverage (4+)?
